---
title: "RStudio Basics"

output: pdf_document
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \fancyhead[L]{}

---

## 1

Load the data set and get the names of all of the variables included. Identify two interesting variables.

```{r}
#load data
load("data/countries2.rda")

#get names of the variables
names(countries2)
```

This data set contains aggregate variables and data about the economy and regarding population indexes.Some of the variables are numerical, while others are categorical.

One potential dependent variable to study could be "kid_mort", or infant's mortality.

```{r}
summary(countries2$kid_mort)
```

We could study infant's mortality as related to the independent variable "health_exp", or expenditures for health-related costs.

```{r}
summary(countries2$health_exp)
```

## 2

List the number of all the variables.

```{r}
#View(countries2)
dim(countries2)
```

The numbers reflect the rows and columns. Therefore the data set 'countries2' contains 195 countries and 49 variables.

## 3

Create a new object, calculated as the percent in the state who approve of the Biden’s performance MINUS the percent in the state who disapprove of Biden’s performance. Sort the data set by the new variable and list the six highest and lowest states.

```{r}
library(readxl)
Approve21 <- read_excel("data/Approve21.xlsx") 
#View(Approve21)

#calculate proportion of approval
Approve21$net_approve<-(Approve21$Approve)-(Approve21$Disapprove)

#Sort the data set by ```Approve21$net_approve``` and list the six highest and lowest states.
Approve21<-Approve21[order(Approve21$net_approve),]

#six highest
head(Approve21)
```

The six highest states are known to be very conservative, and they are primarily southern states. In this region, people are known to prefer Republican candidates over time.

```{r}
#six lowest
tail(Approve21)
```

The six lowest states are primarily coast states and northern states, which are known to overwhelmingly vote liberal and will have potentially favored President Biden during the elections.

## 4

Produce a histogram of a variable and describe what you see.

```{r}
#create histogram
hist(Approve21$net_approve)
#create labels
hist(Approve21$net_approve,
     main="Percentage Difference in Biden's Approval, by State", #Graph title 
     ylab="Number of States", #vertical axis label
     xlab="% Approval") #Horizontal axis label
```

The histogram highlights two distinct sets of states. These either have a 10% to 20% of people that support President Biden, or they have a 10% to 20% of people who disapprove of him. This means that there are some states, those with a negative percentage difference, who will lean toward having a majority of Republican voters, while those with a positive percentage difference will have a majority of Democrat voters. This histogram denotes the political polarization that American politics is currently living in. People either moderately approve or disapprove of a candidate, but no political candidate captures an overwhelming amount of support across the board.