---
title: "Lab 1: Bias in Self-reported Turnout"
author: "Lisa Gandolfi, Tate Portell, Teresa Navarrete, Jean Blakeman"
output:
  pdf_document: default
  html_document: default
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \fancyhead[L]{}
  - \fancyhead[C]{This material should not be shared beyond those who are enrolled in this class}
---

Surveys are frequently used to measure political behavior such as
voter turnout, but some researchers are concerned about the accuracy
of self-reports.  In particular, they worry about possible *social
desirability bias* where in post-election surveys, respondents who did
not vote in an election lie about not having voted because they may
feel that they should have voted.  Is such a bias present in the
American National Election Studies (ANES)?  The ANES is a nation-wide
survey that has been conducted for every election since 1948.  The
ANES conducts face-to-face interviews with a nationally representative
sample of adults.  The table below displays the names and descriptions
of variables in the `turnout.csv` data file.

--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `year`               Election year
 
 `ANES`               ANES estimated turnout (percentage)
 
 `VEP`                Voting Eligible Population (in thousands) 
 
 `VAP`                Voting Age Population (in thousands) 
 
 `total`              Total ballots cast for highest office (in thousands) 
 
 `felons`             Total ineligible felons (in thousands) 
 
 `noncitizens`        Total non-citizens (in thousands) 
 
 `overseas`           Total eligible overseas voters (in thousands)
 
 `osvoters`           Total ballots counted by overseas voters (in thousands)
--------------------------------------------------------------------------------


## Question 1

Load the data into R and check the dimensions of the data. Hint: look in the 'utils' package to learn how to import a .csv file! Use the default settings.

  Then, obtain a summary of the data.  How many observations are
  there?  What is the range of years covered in this data set? 
  
```{r}
library(utils)
turnout <- read.csv("data/turnout.csv")
dim(turnout)
summary(turnout)
```

The dataset turnout contains 14 observations and 9 variables. In addition, the range of years goes from 1980 to 2008.

## Question 2

Calculate the turnout rate based on the voting age population or
  VAP. Note that for this data set, we must add the total number of
  eligible overseas voters since the *VAP* variable does not
  include these individuals in the count. Next, calculate the turnout
  rate using the voting eligible population or VEP.  What difference
  do you observe? Hint: you will want to look at the results across all years. 
  
  Note: to calculate the turnout rate, the numerator is the total number of ballots cast. The denominator is VAP (or VEP). 
  
```{r}
turnout$VAP_ov <- (turnout$VAP + turnout$overseas)
turnout$VAP_total<-(turnout$total/turnout$VAP_ov)

turnout$VEP_ov <- (turnout$VEP + turnout$overseas)
turnout$VEP_total <- (turnout$total/turnout$VEP_ov)
```

The turnout rate for VAP including overseas population is smaller than the turnout rate for VEP and overseas population. This might mean that when felons and non-citizens are included, the voter turnout slightly decreases, holding all else constant.
  
## Question 3

Compute the difference between VAP and ANES estimates of turnout
  rate.  How big is the difference on average?  What is the range of
  the difference?  Conduct the same comparison for the VEP and ANES
  estimates of voter turnout.  Briefly comment on the results.

```{r}
library(descr)
library(DescTools)
library(Hmisc)

#using VAP_total as our turnout rate estimation
turnout$VAP_total_percent <- (turnout$VAP_total*100)
summary(turnout)

turnout$VEP_total_percent <- (turnout$VEP_total*100)
summary(turnout)

```

The difference between VAP_total and ANES turnout rate is on average about 20%, with ANES being the higher value. The range of difference spans from 12.17% of difference to 22.33% of difference in turnout. The difference between VEP_total and ANES turnout rate is on average about 19%, very similar to the difference between VAP and ANES. VEP difference is on average one point percenatge smaller than the VAP difference.